---
title: "P4 — Design-aware main estimate: step-by-step reproducible code"
subtitle: "SPPH 604 · demonstration paper (Kueh et al., BMJ Open 2026;16:e113719)"
format:
  html:
    embed-resources: true
    toc: true
    code-tools: true
    df-print: kable
execute:
  warning: false
  message: false
---

> Replicates the paper's **unweighted** Cox model, then re-estimates it with the
> NHANES **fasting-subsample survey design** (weights + strata + clusters).
>
> **The rule this file exists to demonstrate:** build the survey design on the
> *whole sampled frame*, then reach your analysis population with
> `subset(design, ...)`. Never filter the data frame and then call `svydesign()`
> on the remains.

```{r setup}
source("../_shared/paths.R")   # find_repro(), emit_results(), ci(), fmtn(), endash()
repro <- find_repro()
suppressPackageStartupMessages({library(dplyr); library(survival); library(survey)})
options(survey.lonely.psu = "adjust")

frame <- readRDS(file.path(repro, "data/derived/merged_all.rds"))   # 59,842 records
d     <- readRDS(file.path(repro, "data/derived/masld_analytic.rds"))
frame$in_analysis <- frame$SEQN %in% d$SEQN
frame$group <- factor(frame$group, levels = c("I","II","III","IV"))

hr <- function(m,g){s<-summary(m); sprintf("%.2f (%.2f-%.2f)",
        s$conf.int[g,1], s$conf.int[g,"lower .95"], s$conf.int[g,"upper .95"])}
psu <- function(df) length(unique(paste(df$SDMVSTRA, df$SDMVPSU)))
```

## Step 1 — Replicate the paper (unweighted Cox)

Two row sets, labelled, because they are not interchangeable. The **full analytic
file** (n = 6,371) is what reproduces the paper. The **locked domain** (n = 6,048, those
with a valid fasting weight) is what P1 locks and what P2, P3 and P5 analyse — and the
only set on which a design-aware estimate is possible.

```{r replicate}
dom0 <- d[!is.na(d$WTSAF2YR) & d$WTSAF2YR > 0, ]
mu_full <- coxph(Surv(time_yr,dead) ~ group, data=d)
mu_dom  <- coxph(Surv(time_yr,dead) ~ group, data=dom0)
cat("Unweighted, full analytic file  n =", nrow(d),
    ": IV", hr(mu_full,"groupIV"), " <- reproduces the paper
")
cat("Unweighted, locked domain       n =", nrow(dom0),
    ": IV", hr(mu_dom,"groupIV"),  " <- the rows P2/P3/P5 use
")
```

The difference between them is the records without a valid fasting weight, not a
modelling choice. Everything below uses the locked domain.

## Step 2 — Improve: the survey design, built *before* restricting

FLI needs *fasting* triglycerides, so the binding constraint is the fasting
subsample and the weight is the pooled fasting weight `WTSAF2YR / 6` (6 cycles).

The design is constructed on **everyone with a valid fasting weight** — not on the
MASLD cohort — and the analysis population is then reached with `subset()`. This is
a **domain (subpopulation) analysis**: `subset.survey.design` retains the original
strata and PSU structure so the variance is estimated against the design that
actually produced the sample.

```{r design}
fast <- frame %>% filter(!is.na(WTSAF2YR) & WTSAF2YR > 0) %>% mutate(wt = WTSAF2YR/6)
des_full <- svydesign(ids=~SDMVPSU, strata=~SDMVSTRA, weights=~wt, data=fast, nest=TRUE)

des <- subset(des_full, in_analysis)      # <- the domain. NOT a new svydesign().

dom <- fast %>% filter(in_analysis)
cat("design frame: n =", nrow(fast), " PSUs =", psu(fast),
    " strata =", length(unique(fast$SDMVSTRA)), "\n")
cat("analysis domain: n =", nrow(dom),
    " weighted population =", format(round(sum(dom$wt)), big.mark=","), "\n")
```

```{r weighted}
mu <- coxph(Surv(time_yr,dead) ~ group, data=dom)          # unweighted, same rows
mw <- svycoxph(Surv(time_yr,dead) ~ group, design=des)
cat("Unweighted (domain): IV", hr(mu,"groupIV"), "\n")
cat("Weighted:  II", hr(mw,"groupII"), " III", hr(mw,"groupIII"),
    " IV", hr(mw,"groupIV"), "\n")
```

## Step 3 — Direct contrasts the paper did not report (weighted)

Change the reference level **inside the existing design** with `update()`. Do not
rebuild `svydesign()` — that would discard the frame a second time.

```{r contrasts}
des3 <- update(des, group = relevel(group, "III"))
des2 <- update(des, group = relevel(group, "II"))
mw3 <- svycoxph(Surv(time_yr,dead) ~ group, design=des3)
mw2 <- svycoxph(Surv(time_yr,dead) ~ group, design=des2)
cat("Weighted IV vs III:", hr(mw3,"groupIV"), "   IV vs II:", hr(mw2,"groupIV"), "\n")
```

## Step 3b — Separating adjustment from design on one row set

The ledger 15.1 -> 2.9 -> 12.4 mixes crude and adjusted models on different rows, so it
is not a sequence of successive corrections. Hold the rows fixed and the two effects
separate cleanly.

```{r matrix}
COV <- c("age","female","dyslip","htn","t2dm","ckd","mi","cancer",
         "on_htn_med","on_dm_med","on_lipid_med")
fast$cc <- fast$in_analysis & complete.cases(fast[,COV])
dcc <- fast[fast$cc, ]
f0 <- Surv(time_yr,dead) ~ group
f1 <- as.formula(paste("Surv(time_yr,dead) ~ group +", paste(COV, collapse=" + ")))
gg <- function(f){s<-summary(f); sprintf('%.2f (%.2f-%.2f)',
        s$conf.int['groupIV',1], s$conf.int['groupIV',3], s$conf.int['groupIV',4])}
rl <- function(x,r){x$group <- relevel(x$group, r); x}
dR <- function(r) subset(svydesign(ids=~SDMVPSU, strata=~SDMVSTRA, weights=~wt,
                    data=rl(fast,r), nest=TRUE), cc)
mat <- data.frame(
  cell = c('unweighted crude','unweighted Model 1','weighted crude','weighted Model 1'),
  IV_vs_I = c(gg(coxph(f0,data=dcc)), gg(coxph(f1,data=dcc)),
              gg(svycoxph(f0,design=dR('I'))), gg(svycoxph(f1,design=dR('I')))),
  IV_vs_III = c(gg(coxph(f0,data=rl(dcc,'III'))), gg(coxph(f1,data=rl(dcc,'III'))),
                gg(svycoxph(f0,design=dR('III'))), gg(svycoxph(f1,design=dR('III')))))
mat
```

Read **down** a column for adjustment, **across** a row for design. On the locked
IV-vs-III estimand adjustment dominates and design changes no conclusion; on the
replication IV-vs-I estimand design changes whether the adjusted interval excludes 1.

## Step 4 — Why the rule matters (and exactly when it bites)

For *this* main estimate both approaches give the same answer, because the MASLD
domain happens to contain every PSU and stratum in the fasting frame. Pre-filtering
loses nothing here. That is luck, not method.

```{r why}
show <- function(lbl, df) cat(sprintf("%-32s n=%5d  PSUs=%4d  strata=%3d
", lbl, nrow(df), psu(df), length(unique(df$SDMVSTRA))))
show("full fasting frame", fast)
show("MASLD analysis domain", dom)
show("  Group I only", dom %>% filter(group=="I"))
show("  Group I, women only", dom %>% filter(group=="I", female==1))
```

The domain keeps all 184 PSUs. A subgroup does not: Group I women sit in ~51 PSUs
spread over 44 strata, which leaves **strata containing a single cluster**. That is
where pre-filtering stops being harmless.

```{r lonely}
d2 <- dom %>% filter(group=="I", female==1)
for (opt in c("fail","adjust")) {
  options(survey.lonely.psu = opt)
  dfull <- svydesign(ids=~SDMVPSU, strata=~SDMVSTRA, weights=~wt, data=fast, nest=TRUE)
  pre <- tryCatch(sprintf("%.4f", SE(svymean(~age, svydesign(ids=~SDMVPSU,
           strata=~SDMVSTRA, weights=~wt, data=d2, nest=TRUE)))),
         error=function(e) paste0("ERROR: ", conditionMessage(e)))
  sub <- sprintf("%.4f", SE(svymean(~age,
           subset(dfull, in_analysis & group=="I" & female==1))))
  cat(sprintf("lonely.psu=%-7s  pre-filtered: %-45s  subset(): %s
", opt, pre, sub))
}
options(survey.lonely.psu = "adjust")
```

Read that table carefully, because the lesson is subtler than "pre-filtering gives
the wrong number":

- Under R's **default** (`lonely.psu = "fail"`), pre-filtering does not give a wrong
  answer — it gives **no answer**. The design you handed `svydesign()` has strata
  with one cluster and no variance is estimable.
- Setting `lonely.psu = "adjust"` makes it run, and here it returns the same SE. But
  you have silenced a symptom of having thrown the design away, and you are now
  relying on an approximation instead of the stratum structure you actually had.
- `subset()` needs no rescue. It works under either setting, because the strata and
  PSUs were never discarded.

So the rule is not "pre-filtering biases your estimate." It is: **pre-filtering can
create lonely-PSU strata that would not arise under proper domain analysis, making the
variance estimate depend on the chosen lonely-PSU fallback.** Restricting the
6,048-person MASLD cohort did nothing at all, because it retained every PSU and
stratum. The failure appears once restriction starts dropping them. Sometimes the
fallback agrees; you will not know which time you are in.

Note what this separates: **a correct procedure is not the same as a procedure that
changes today's answer.** The wrong implementation returns an identical HR and CI in
the main analysis and is still wrong, because it is not guaranteed to preserve the
variance structure for narrower domains.

**Your own paper's subgroups will be sparser than this one.** Build the design once,
then subset.

**Interpretation.** Weighting attenuates the Group IV estimate (≈ 17 → ≈ 12) and
widens the CI — that is the design effect. Group IV stays worst even against the
other high-risk groups (IV vs III ≈ 3.1; IV vs II ≈ 2.3).

**On which weight.** CDC's rule is the *least common denominator*: when an analysis
draws on variables measured in different components, use the weight belonging to
the **smallest** applicable subsample. FLI requires fasting triglycerides, so the
fasting weight governs, even though GGT is measured on the larger MEC sample. This
is a rule, not a judgment call — the MEC weight would weight to a population in
which the exposure cannot be constructed.

## Results for M1

```{r emit}
cell <- function(row, col) endash(mat[[col]][mat$cell == row])
emit_results("P4", list(
  uw_crude_i    = cell("unweighted crude",   "IV_vs_I"),
  uw_crude_iii  = cell("unweighted crude",   "IV_vs_III"),
  uw_m1_i       = cell("unweighted Model 1", "IV_vs_I"),
  uw_m1_iii     = cell("unweighted Model 1", "IV_vs_III"),
  w_crude_i     = cell("weighted crude",     "IV_vs_I"),
  w_crude_iii   = cell("weighted crude",     "IV_vs_III"),
  w_m1_i        = cell("weighted Model 1",   "IV_vs_I"),
  w_m1_iii      = cell("weighted Model 1",   "IV_vs_III"),
  n_matched     = fmtn(nrow(dcc)),
  psu_full      = fmtn(psu(fast)),
  psu_domain    = fmtn(psu(dom))
), list(matrix = mat))
```
