---
title: "M3 — Defense evidence pack: reproducible numbers behind the Q&A"
subtitle: "SPPH 604 · demonstration paper (Kueh et al., BMJ Open 2026;16:e113719)"
format:
  html:
    embed-resources: true
    toc: true
    code-tools: true
    df-print: kable
execute:
  warning: false
  message: false
---

> The individual defense is not a slide-reading exercise — every number you may be
> asked to justify is re-derived here so you can answer from evidence, live.

```{r setup}
source("../_shared/paths.R")   # find_repro(), emit_results(), ci(), fmtn(), endash()
repro <- find_repro()
suppressPackageStartupMessages({library(dplyr); library(survival); library(survey); library(mice)})
options(survey.lonely.psu = "adjust")
d <- readRDS(file.path(repro, "data/derived/masld_analytic.rds"))
hr <- function(m,g){s<-summary(m); sprintf("%.2f (%.2f-%.2f)",
        s$conf.int[g,1], s$conf.int[g,"lower .95"], s$conf.int[g,"upper .95"])}
```

## Q (P1): the cohort/N discrepancy

```{r q1}
cat("Our N =", nrow(d), "vs paper 6300; complete-case =",
    nrow(filter(d,!is.na(cancer),!is.na(stroke),!is.na(hf),!is.na(LBXSCR))),
    "; missing creatinine =", sum(is.na(d$LBXSCR)), "\n")
```

## Q (P3): is the sex "interaction" real, or baseline risk?

```{r q3}
cov <- "age + dyslip + htn + t2dm + ckd + mi + cancer + on_htn_med + on_dm_med + on_lipid_med"
adj_p <- anova(coxph(as.formula(paste("Surv(time_yr,dead)~group+female+",cov)),d),
               coxph(as.formula(paste("Surv(time_yr,dead)~group*female+",cov)),d))[2,"Pr(>|Chi|)"]
cat("Adjusted group x sex interaction p =", round(adj_p,3),
    "-> no evidence of MULTIPLICATIVE interaction on this multi-df test.\n")
cat("   Not evidence of homogeneity: the test is underpowered, and it is not the locked\n")
cat("   1-df IV-vs-III contrast (see ../P3_effect_modification/P3_code.qmd).",
    "Additive scale untested.\n")
```

## Q (P4): which weight, and what does it do?

```{r q4}
# Week 6 cardinal rule: build the design on the FULL fasting frame, then subset the design
# object. Filtering rows first and re-declaring svydesign() on the survivors throws away
# the strata/PSU information the non-selected units contribute to the variance.
fr <- readRDS(file.path(repro,"data/derived/merged_all.rds")) %>%
  filter(!is.na(WTSAF2YR) & WTSAF2YR>0) %>% mutate(wt=WTSAF2YR/6)
fr$group       <- factor(fr$group, levels=c("I","II","III","IV"))
fr$in_analysis <- fr$SEQN %in% d$SEQN
des_full <- svydesign(ids=~SDMVPSU, strata=~SDMVSTRA, weights=~wt, data=fr, nest=TRUE)
des <- subset(des_full, in_analysis)      # domain analysis, not a re-declared design
dd  <- fr[fr$in_analysis, ]
cat("Group IV: unweighted", hr(coxph(Surv(time_yr,dead)~group,dd),"groupIV"),
    "-> fasting-weighted", hr(svycoxph(Surv(time_yr,dead)~group,design=des),"groupIV"), "\n")
```

## Q (self-initiated): could incomplete cancer status explain the locked result?

> **No reviewer asked for this.** We ran it because cancer status is the only Model-1
> covariate with any missingness, and a result that rests on dropping people is worth
> testing before we defend it. The adjustment set is the paper's Model 1 exactly — we did
> **not** widen it to manufacture a comparison. The contrast is our locked one, IV vs III,
> design-aware, so the sensitivity analysis interrogates the estimate we actually defend.

```{r q5}
m <- readRDS(file.path(repro, "data/derived/merged_all.rds"))
m$in_analysis <- m$SEQN %in% d$SEQN
m$group <- factor(m$group, levels = c("I", "II", "III", "IV"))
fast <- m %>% filter(!is.na(WTSAF2YR), WTSAF2YR > 0) %>% mutate(wt = WTSAF2YR / 6)

COV <- c("age","female","dyslip","htn","t2dm","ckd","mi","cancer",
         "on_htn_med","on_dm_med","on_lipid_med")            # the paper's Model 1. Nothing else.
f1  <- as.formula(paste("Surv(time_yr,dead) ~ group +", paste(COV, collapse = " + ")))

# locked contrast: Group III is the reference, so groupIV is IV-vs-III
rl3  <- function(x) { x$group <- relevel(x$group, "III"); x }
desL <- function(fk) subset(svydesign(ids = ~SDMVPSU, strata = ~SDMVSTRA, weights = ~wt,
                                      data = rl3(fk), nest = TRUE), in_analysis)
gv   <- function(f) { s <- summary(f); c(s$conf.int["groupIV", 1],
                                         s$conf.int["groupIV", 3], s$conf.int["groupIV", 4]) }

fitL <- function(val, lbl) {
  fk <- fast
  if (!is.na(val)) fk$cancer[fk$in_analysis & is.na(fk$cancer)] <- val
  g <- gv(svycoxph(f1, design = desL(fk)))
  data.frame(scenario = lbl, HR = round(g[1], 2), lo = round(g[2], 2), hi = round(g[3], 2))
}
rbind(fitL(NA, "complete case"),
      fitL(0,  "all missing assigned NO cancer"),
      fitL(1,  "all missing assigned CANCER"))
```

```{r q5mi}
# MI with the outcome in the imputation model (Week 7 / White & Royston): keep the event
# indicator and a Nelson-Aalen cumulative hazard, exclude only raw follow-up time.
dom  <- fast %>% filter(in_analysis)
idat <- dom[, c(COV, "group", "dead", "time_yr", "wt", "SDMVSTRA")]
idat$na_cumhaz <- mice::nelsonaalen(idat, time_yr, dead)
ini <- mice(idat, maxit = 0, printFlag = FALSE); pr <- ini$predictorMatrix
pr[, "time_yr"] <- 0
pr[c("time_yr","dead","na_cumhaz","wt","SDMVSTRA"), ] <- 0
imp <- mice(idat, m = 10, predictorMatrix = pr, printFlag = FALSE, seed = 604)

fits <- lapply(1:10, function(i) {
  fk <- fast; fk[fk$in_analysis, COV] <- complete(imp, i)[, COV]
  svycoxph(f1, design = desL(fk))
})
po <- summary(mitools::MIcombine(fits))["groupIV", ]
cat(sprintf("MI (m=10), locked contrast: IV vs III = %.2f (%.2f-%.2f)\n",
            exp(po[["results"]]), exp(po[["(lower"]]), exp(po[["upper)"]])))
cat("Complete case, both extreme assignments and MI agree to the displayed precision.\n")
cat("That is the finding. It is evidence that incomplete cancer status does not explain\n")
cat("the locked estimate -- it says nothing about who the estimate generalises to.\n")
```

## Q (reviewer-prompted): what exactly is the selection, and what does it bound?

> Our M2 reviewers named selection into the cohort as the dominant threat. Their slide
> treats it as one mechanism. It is **two**, and they bound different things, so we
> separate them before answering.

```{r q6}
adults <- m %>% filter(age >= 18)
fast_a <- adults %>% filter(!is.na(WTSAF2YR), WTSAF2YR > 0)
fli_ok <- fast_a %>% filter(!is.na(fli))
cat(sprintf("adults 18+                                  %6d\n", nrow(adults)))
cat(sprintf("  (1) sampled into the fasting subsample    %6d\n", nrow(fast_a)))
cat(sprintf("  (2) of those, FLI computable              %6d\n", nrow(fli_ok)))
cat(sprintf("      lost at step (2)                      %6d\n", nrow(fast_a) - nrow(fli_ok)))
```

- **(1) is selection by design.** NHANES chose who was asked to fast. It is known,
  planned, and it is what `WTSAF2YR` exists to account for — which is why we weight by it
  and subset the design rather than the rows. Under the survey design this step is
  *handled*, not merely acknowledged.
- **(2) is item-level availability.** Among people who were sampled to fast, FLI still
  needs triglycerides, GGT, BMI and waist all present. Nothing in the design weights
  accounts for that, and whether it is ignorable depends on why those measurements are
  absent — which we cannot verify from the public files.

**What each one bounds.** (1) bounds the *target population*: our estimate speaks to
fasting-sampled US adults, and the weights carry it there. (2) is a potential selection
bias *within* that population, and it is the one that could distort the contrast itself.

**Our answer to the reviewers.** We accept the distinction matters and have adopted it.
We do not accept that stability of the estimate answers it: the missing-data work above
concerns incomplete **cancer status among people already in the cohort**, and tells us
nothing about the adults who never reach the cohort at all -- the two steps counted above. Those are different questions and we
now report them separately.

## Results manifest — the chain's single source of truth

> M2, M3 and M4 are plain markdown and are **not** run through the deck generator, so any
> number typed into them could drift from the code. This chunk emits the canonical values;
> the build fails if any of those artifacts states a locked-contrast hazard
> ratio that is not the one recorded here.

```{r emit}
cc <- fitL(NA, "cc")
mi_hr <- exp(po[["results"]]); mi_lo <- exp(po[["(lower"]]); mi_hi <- exp(po[["upper)"]])
emit_results("chain", list(
  locked_hr      = sprintf("%.2f", cc$HR),
  locked_ci      = sprintf("%.2f–%.2f", cc$lo, cc$hi),
  locked_full    = sprintf("%.2f (%.2f–%.2f)", cc$HR, cc$lo, cc$hi),
  locked_mi_full = sprintf("%.2f (%.2f–%.2f)", mi_hr, mi_lo, mi_hi),
  n_adults       = format(nrow(adults), big.mark = ","),
  n_fasting      = format(nrow(fast_a), big.mark = ","),
  n_fli          = format(nrow(fli_ok), big.mark = ","),
  n_lost_step2   = format(nrow(fast_a) - nrow(fli_ok), big.mark = ",")
))
```
