---
title: "M4 — Reanalysis letter: master reproducibility file"
subtitle: "SPPH 604 · demonstration paper (Kueh et al., BMJ Open 2026;16:e113719)"
format:
  html:
    embed-resources: true
    toc: true
    code-tools: true
    df-print: kable
execute:
  warning: false
  message: false
---

> Reproduces every number stated in the letter. The replication figures are
> derived live here, except the Model 1 hazard ratio, which comes from the
> reproduction's `../../reproduction/output/tables/table2_reproduced.csv`. The locked-contrast estimates
> (Group IV vs III, the paper's Model 1, design-weighted) and the funnel counts are
> emitted by `../P5_missing_data/P5_code.qmd` and `../M3_defense/M3_defense_evidence.qmd`,
> and are read below rather than retyped.

```{r setup}
source("../_shared/paths.R")   # find_repro(), emit_results(), ci(), fmtn(), endash()
repro <- find_repro()
suppressPackageStartupMessages({library(dplyr); library(survival)})
d <- readRDS(file.path(repro, "data/derived/masld_analytic.rds"))
hr <- function(m,g){s<-summary(m); sprintf("%.2f (%.2f-%.2f)",
        s$conf.int[g,1], s$conf.int[g,"lower .95"], s$conf.int[g,"upper .95"])}
```

## Claim 1 — the cohort and headline results, against the paper's

```{r reproduce}
cat("N =", nrow(d), "(paper 6300);  groups",
    paste(as.integer(table(d$group)), collapse="/"), "(386/4541/769/604)\n")
cat("Deaths =", sum(d$dead,na.rm=TRUE), "(585);  by group",
    paste(as.integer(tapply(d$dead,d$group,sum,na.rm=TRUE)), collapse="/"), "(5/388/74/118)\n")
cov1 <- c("age","female","dyslip","htn","t2dm","ckd","mi","cancer","on_htn_med","on_dm_med","on_lipid_med")
d$smk <- ifelse(d$SMQ020==1 & d$SMQ040%in%c(1,2),1L,ifelse(is.na(d$SMQ020),NA,0L))
u  <- coxph(Surv(time_yr,dead)~group, d)
m2 <- coxph(Surv(time_yr,dead)~group+age+female+dyslip+htn+t2dm+ckd+mi+cancer+
            on_htn_med+on_dm_med+on_lipid_med+smk+sedentary_min, d)
cat("Group IV all-cause: unadjusted", hr(u,"groupIV"), "(paper 15.13);  Model 2",
    hr(m2,"groupIV"), "(paper 2.89)\n")
```

## Claim 2/4 — the locked contrast, and its missing-data sensitivity

```{r improved}
# Read, not retyped: P5.json is emitted by ../P5_missing_data/P5_code.qmd and chain.json
# by ../M3_defense/M3_defense_evidence.qmd. All rows: Group IV vs III, the paper's
# Model 1, design-weighted (WTSAF2YR/6, strata, PSU).
p5 <- read_results("P5")$facts; ch <- read_results("chain")$facts
data.frame(
  Analysis = c("complete case (the defended estimate)", "all missing assigned no cancer",
               "all missing assigned cancer", "MI (m = 10), outcome in the imputation model"),
  Rows = c(p5$n_cc, p5$n_full, p5$n_full, p5$n_full),
  `HR (95% CI)` = c(ch$locked_full, p5$locked_no, p5$locked_yes, ch$locked_mi_full),
  Source = c("M3_defense_evidence.qmd", "P5_code.qmd", "P5_code.qmd", "M3_defense_evidence.qmd"),
  check.names = FALSE)
cat("Funnel: adults 18+", ch$n_adults, "; sampled into the fasting subsample", ch$n_fasting,
    "; of those, FLI computable", ch$n_fli, "\n")
```

## Claim 3 — published-vs-reproduced discrepancies we flag

```{r errors}
pir <- table(d$pir_cat); den <- sum(pir)
cat("PIR counts imply High", round(100*pir['High']/den,1), "/ Low", round(100*pir['Low']/den,1),
    "; paper prints 23.0 / 22.1 -> printed percentages disagree with the counts.\n")
cat("BMI: non-obese group medians", round(median(d$bmi[d$group=='III']),1), "/",
    round(median(d$bmi[d$group=='IV']),1), "-> incompatible with the '<25' header (Methods cut = 30).\n")
```

*Data availability:* all inputs are public (NHANES 2007–2018; NCHS linked mortality).
The full pipeline is in `../../reproduction/`.
