---
title: "P3 — Effect modification (interaction): step-by-step reproducible code"
subtitle: "SPPH 604 · demonstration paper (Kueh et al., BMJ Open 2026;16:e113719)"
format:
  html:
    embed-resources: true
    toc: true
    code-tools: true
    df-print: kable
execute:
  warning: false
  message: false
---

> Tests whether the phenotype → mortality association is modified by **sex**.
> The crude sex-stratified HRs look dramatically different; the targeted adjusted
> test is centred near 1 but far too imprecise to settle the question. Reporting
> that imprecision honestly is the lesson of this increment.

```{r setup}
source("../_shared/paths.R")   # find_repro(), emit_results(), ci(), fmtn(), endash()
repro <- find_repro()
suppressPackageStartupMessages({library(dplyr); library(survival)})
## The cohort is the one P1 locks: the MASLD domain of the fasting subsample.
## All five increments use these rows so P4 can weight exactly what P2-P3 analysed.
.a <- readRDS(file.path(repro, "data/derived/masld_analytic.rds"))
.m <- readRDS(file.path(repro, "data/derived/merged_all.rds"))
.m$in_analysis <- .m$SEQN %in% .a$SEQN
d  <- .m[.m$in_analysis & !is.na(.m$WTSAF2YR) & .m$WTSAF2YR > 0, ]
d$group <- factor(d$group, levels = c("I","II","III","IV"))
cat("cohort n =", nrow(d), "(P1-locked domain)
")
hr <- function(m,g){s<-summary(m); sprintf("%.2f (%.2f-%.2f)",
        s$conf.int[g,1], s$conf.int[g,"lower .95"], s$conf.int[g,"upper .95"])}
```

Rationale: sex is entangled with the exposure — Group II is 56% female, Group I only 15% —
so modification by sex is worth probing.

## Sex-stratified group hazard ratios (all-cause, vs Group I)

```{r stratified}
strat <- sapply(c(0, 1), function(s) {
  ms <- coxph(Surv(time_yr,dead) ~ group, data = filter(d, female == s))
  c(II = hr(ms,"groupII"), III = hr(ms,"groupIII"), IV = hr(ms,"groupIV"))
})
colnames(strat) <- c("Men", "Women")
pt <- function(sex) sub(" .*", "", strat["IV", sex])   # point estimate only
t(strat)
```

Men look far higher (IV `r pt("Men")`) than women (IV `r pt("Women")`). But the
reference group has very few events, and men in this cohort are older and sicker — so
is this real modification, or confounding / baseline-risk difference?

*Those two figures are inline references to the table above, not typed. An earlier
draft typed "IV ≈ 28" beside a chunk that printed something else; the fix is to make
the prose read from the object rather than to correct the digits.*

## Formal interaction test — targeted at the locked contrast

Our locked estimand is **IV vs III**. A `group * sex` term across all four groups is a
**multi-df test of whether *any* contrast varies by sex** — a broader question. We
restrict to Groups III and IV so the interaction is **1 df for exactly our contrast**,
and fit crude and adjusted on the **same** complete-case rows.

```{r interaction}
COVns <- c("age","dyslip","htn","t2dm","ckd","mi","cancer",
           "on_htn_med","on_dm_med","on_lipid_med")
s   <- d %>% filter(group %in% c("III","IV")) %>% mutate(IV = as.integer(group=="IV"))
scc <- s[complete.cases(s[,c(COVns,"female")]),]
tst <- function(f){
  full <- coxph(f, data=scc); red <- coxph(update.formula(f, .~.-IV:female), data=scc)
  b <- coef(full)[["IV:female"]]; se <- sqrt(vcov(full)["IV:female","IV:female"])
  data.frame(rHR=sprintf("%.2f (%.2f-%.2f)", exp(b), exp(b-1.96*se), exp(b+1.96*se)),
             p=sprintf("%.3f", anova(red, full)$`Pr(>|Chi|)`[2]))
}
res <- rbind(cbind(Model="crude",   tst(Surv(time_yr,dead) ~ IV*female)),
             cbind(Model="Model 1", tst(as.formula(paste(
                   "Surv(time_yr,dead) ~ IV*female +", paste(COVns, collapse=" + "))))))
cat("rows used:", nrow(scc), " women:", sum(scc$female==1), " men:", sum(scc$female==0), "\n")
res
```

Ratio of hazard ratios is women / men for the IV-vs-III contrast.

## Additive scale — six-year standardized risk difference

Stratum-specific hazard ratios do not show the additive picture; it has to be estimated.
We fit Model 1 within each sex, predict six-year risk for every person in that stratum
under `IV = 0` and `IV = 1`, and average — g-computation, so the contrast is a
population-average risk difference over that stratum's covariate distribution.

```{r additive}
risk6 <- function(dd){
  fit <- coxph(as.formula(paste("Surv(time_yr,dead) ~ IV +", paste(COVns, collapse=" + "))), data=dd)
  r <- sapply(c(0,1), function(v){ z <- dd; z$IV <- v
    mean(1 - summary(survfit(fit, newdata=z), times=6, extend=TRUE)$surv) })
  c(r3=r[1], r4=r[2], rd=r[2]-r[1])
}
set.seed(604)
add <- do.call(rbind, lapply(c(0,1), function(sx){
  dd <- scc %>% filter(female==sx); pt <- risk6(dd)
  bs <- replicate(300, tryCatch(risk6(dd[sample(nrow(dd), replace=TRUE),])[["rd"]],
                                error=function(e) NA_real_))
  ci <- quantile(bs, c(.025,.975), na.rm=TRUE)
  data.frame(Sex=ifelse(sx==1,"women","men"), n=nrow(dd),
             risk_III=round(pt[["r3"]],3), risk_IV=round(pt[["r4"]],3),
             RD=sprintf("%+.3f (%.3f to %.3f)", pt[["rd"]], ci[1], ci[2]))
}))
add
```

Some bootstrap replicates in the female stratum trigger Cox convergence warnings — 359
women with sparse events means some resamples separate — so that interval is wide and
should be read as order-of-magnitude.

**Conclusion, stated at the precision the data afford.** The adjusted ratio of hazard
ratios is 1.00 with an interval running from 0.42 to 2.35 — compatible with the effect in
women being less than half that in men, or more than twice it. Both six-year risk
differences are under one percentage point. So the data show **no evidence of
multiplicative interaction at the available precision**; they do not show that sex is
irrelevant, and they do not identify what generated the crude stratified gap.

**What this file deliberately does not do.** Everything above is **unweighted**, because
P3 precedes the survey-design increment. P4 shows that design materially changes
inference for the main estimate, so whether the design-aware interaction agrees is an
open question — one M1 answers, by applying P4's domain rule to this contrast.

## Results for M1

```{r emit}
pick <- function(col, model) res[[col]][res$Model == model]
emit_results("P3", list(
  rhr_crude      = endash(pick("rHR", "crude")),
  p_crude        = pick("p", "crude"),
  rhr_unweighted = endash(pick("rHR", "Model 1")),
  p_unweighted   = pick("p", "Model 1"),
  n        = fmtn(nrow(scc)),
  n_men    = fmtn(sum(scc$female == 0)),
  n_women  = fmtn(sum(scc$female == 1)),
  rd_men   = add$RD[add$Sex == "men"],
  rd_women = add$RD[add$Sex == "women"]
), list(interaction = res, additive = add))
```
