---
title: "P2 — Confounding diagnostic (Table 1): step-by-step reproducible code"
subtitle: "SPPH 604 · demonstration paper (Kueh et al., BMJ Open 2026;16:e113719)"
format:
  html:
    embed-resources: true
    toc: true
    code-tools: true
    df-print: kable
execute:
  warning: false
  message: false
---

> Reproduces the Table 1 rows we cite, the **PIR income-% swap**, and the
> crude → Model 1 attenuation of the Group IV hazard ratio.

```{r setup}
source("../_shared/paths.R")   # find_repro(), emit_results(), ci(), fmtn(), endash()
repro <- find_repro()
suppressPackageStartupMessages({library(dplyr); library(survival)})
## The cohort is the one P1 locks: the MASLD domain of the fasting subsample.
## All five increments use these rows so P4 can weight exactly what P2-P3 analysed.
.a <- readRDS(file.path(repro, "data/derived/masld_analytic.rds"))
.m <- readRDS(file.path(repro, "data/derived/merged_all.rds"))
.m$in_analysis <- .m$SEQN %in% .a$SEQN
d  <- .m[.m$in_analysis & !is.na(.m$WTSAF2YR) & .m$WTSAF2YR > 0, ]
d$group <- factor(d$group, levels = c("I","II","III","IV"))
cat("cohort n =", nrow(d), "(P1-locked domain)
")
```

## Table 1 — key rows, reproduced vs published

```{r table1}
med_iqr <- function(x) sprintf("%.1f (%.1f-%.1f)", median(x,na.rm=T),
                               quantile(x,.25,na.rm=T), quantile(x,.75,na.rm=T))
mean_sd <- function(x) sprintf("%.1f (%.1f)", mean(x,na.rm=T), sd(x,na.rm=T))
npct    <- function(x){x<-x[!is.na(x)]; sprintf("%d (%.1f)", sum(x==1), 100*mean(x==1))}
data.frame(
  Characteristic = c("Age (mean SD)","Female n(%)","Waist median(IQR)","BMI median(IQR)",
                     "SBP","DBP","Hypertension","T2DM","Dyslipidaemia","CKD","MI",
                     "Heart failure","Cancer"),
  Reproduced = c(mean_sd(d$age), npct(d$female), med_iqr(d$waist), med_iqr(d$bmi),
                 med_iqr(d$sbp), med_iqr(d$dbp), npct(d$htn), npct(d$t2dm),
                 npct(d$dyslip), npct(d$ckd), npct(d$mi), npct(d$hf), npct(d$cancer)),
  Published  = c("51.0 (16.6)","2973 (47.2)","111.0 (104.0-120.0)","33.2 (30.2-37.4)",
                 "124.0 (114.0-135.0)","72.0 (64.7-79.3)","3238 (52.6)","1645 (26.1)",
                 "3612 (57.7)","1142 (18.1)","518 (8.4)","283 (4.6)","624 (10.1)"))
```

Composite definitions used (calibrated to the paper): HTN = self-report **or** BP ≥ 140/90;
T2DM = self-report **or** HbA1c ≥ 6.5; dyslipidaemia = self-report **or** lipid meds **or** LDL ≥ 130;
MI = heart attack **or** CHD; CKD = eGFR < 60 (CKD-EPI 2021) **or** ACR ≥ 30.

## The PIR income-percentage swap (published vs reconstructed)

```{r pir}
pir <- table(d$pir_cat); den <- sum(pir)
data.frame(
  category = c("High","Middle","Low"),
  count = as.integer(pir[c("High","Middle","Low")]),
  count_pct = round(100*as.integer(pir[c("High","Middle","Low")])/den,1),
  paper_printed = c(23.0,54.9,22.1))
```

The counts imply High **22.1** / Low **23.0**, but the paper prints 23.0 / 22.1 —
the high and low percentages are transposed.

## Two estimands, and why the rows must match

The paper references contrasts to **Group I** (obese, low central). Our locked term
question references **Group III** (non-obese, low central), isolating central adiposity
with body-mass category held fixed. Both come from the *same* fitted model; re-levelling
is reparameterization, not reanalysis.

Model 1 loses 109 observations to missing cancer status, so a crude estimate on all rows
and an adjusted estimate on complete cases differ for **two** reasons at once. We fit
both on the same complete-case set and show the all-rows crude separately.

```{r cox}
COV <- c("age","female","dyslip","htn","t2dm","ckd","mi","cancer",
         "on_htn_med","on_dm_med","on_lipid_med")
f0 <- Surv(time_yr,dead) ~ group
f1 <- as.formula(paste("Surv(time_yr,dead) ~ group +", paste(COV, collapse=" + ")))
hr <- function(m,g){s<-summary(m); sprintf("%.2f (%.2f-%.2f)",
        s$conf.int[g,1], s$conf.int[g,"lower .95"], s$conf.int[g,"upper .95"])}
cc  <- complete.cases(d[,COV]); dcc <- d[cc,]
rel <- function(x,ref){x$group <- relevel(x$group, ref); x}

out <- rbind(
  data.frame(Estimand="replication  IV vs I",  Model="crude",   rows=nrow(d),
             HR=hr(coxph(f0, data=d),   "groupIV")),
  data.frame(Estimand="replication  IV vs I",  Model="crude",   rows=nrow(dcc),
             HR=hr(coxph(f0, data=dcc), "groupIV")),
  data.frame(Estimand="replication  IV vs I",  Model="Model 1", rows=nrow(dcc),
             HR=hr(coxph(f1, data=dcc), "groupIV")),
  data.frame(Estimand="locked       IV vs III", Model="crude",   rows=nrow(d),
             HR=hr(coxph(f0, data=rel(d,"III")),   "groupIV")),
  data.frame(Estimand="locked       IV vs III", Model="crude",   rows=nrow(dcc),
             HR=hr(coxph(f0, data=rel(dcc,"III")), "groupIV")),
  data.frame(Estimand="locked       IV vs III", Model="Model 1", rows=nrow(dcc),
             HR=hr(coxph(f1, data=rel(dcc,"III")), "groupIV")))
out
```

The two crude rows within each estimand differ only by the 109 dropped observations, and
they barely move. **The row restriction accounts for essentially none of the change**, so
what remains is attributable to adjustment — which is the only reason we are entitled to
say so.

On the locked contrast the crude excess does not survive adjustment. On the paper's own
reference it does. Both are properties of one model, answering two different questions.

## Results for M1

```{r emit}
emit_results("P2", list(
  n_domain   = fmtn(nrow(d)),
  n_complete = fmtn(nrow(dcc)),
  n_dropped  = fmtn(nrow(d) - nrow(dcc))
), list(estimands = out))
```
