---
title: "M1 — Assembly notebook: every figure on the deck, and where it came from"
subtitle: "SPPH 604 · demonstration paper (Kueh et al., BMJ Open 2026;16:e113719)"
format:
  html:
    embed-resources: true
    toc: true
    code-tools: true
    df-print: kable
execute:
  warning: false
  message: false
---

> **This file assembles; it does not re-implement.**
>
> Each increment ends by writing a result object to `../_results/`, containing a
> flat set of **facts**: named, already-formatted values. `M1_code.qmd` reads
> those, adds the facts from its own two analyses, and writes `M1.json` and
> `paper.json`.
>
> **The deck contains no typed numbers.** `M1_replication_deck.md` cites facts by
> name — `{{ P4.facts.w_m1_iii }}` — and the deck generator substitutes them at
> build time. A misspelled or stale reference does not resolve, and an unresolved
> reference fails the build. This replaces the earlier self-test, which only asked
> whether a numeric string occurred *somewhere* in both the result object and the
> deck; that test could not tell a real citation from a coincidental match.
>
> **What this does and does not guarantee.** It guarantees that every number
> printed on a slide was produced by the named field of a named result object, and
> that no number was retyped. It does **not** guarantee that the prose around a
> number describes it correctly. Numeric drift is now structurally prevented;
> interpretive drift is still a matter of reading the slides.
>
> **M1 runs exactly two analyses of its own:**
>
> 1. **The replication table** — the paper's Table 2 on the full analytic file.
>    Replication is M1's task; no increment does it.
> 2. **The design-aware interaction.** P3 states that it is unweighted and that
>    "whether the design-aware interaction gives the same qualitative answer is an
>    open question, and one M1 should check rather than assume." This file checks
>    it, applying P4's domain rule to P3's contrast.

```{r setup}
source("../_shared/paths.R")   # find_repro(), find_paper(), emit_results(), ci(), fmtn()
repro <- find_repro()
suppressPackageStartupMessages({
  library(dplyr); library(survival); library(survey)})
options(survey.lonely.psu = "adjust")

R <- setNames(lapply(sprintf("P%d", 1:5), read_results), sprintf("P%d", 1:5))
cat("assembled from:\n")
for (n in names(R)) cat(sprintf("  %-4s %2d facts\n", n, length(R[[n]]$facts)))
```

## Figures quoted from the article, checked against the article

Slides 2, 4, 7 and 8 cite the paper. Those numbers are not ours to reproduce, so
each is transcribed once here **with the string that must be found in the PDF**.
If a check fails, the render fails: we do not want a quoted figure that the source
does not contain.

**Three tiers, and the table below names the tier for every fact.** This matters,
because a check that only asks "does this string occur?" can never fail on the value
we actually display:

- **exact** — the displayed value *is* the string sought in the PDF, so presence and
  correctness are the same test.
- **derived** — the displayed value is *reconstructed* from the PDF and compared to
  what we transcribed. A typo in our display fails the build even though the
  supporting strings are all present. Aggregates and reformatted intervals are here.
- **presence** — only the supporting strings are confirmed. Prose quotations, and
  table cells whose row the article's two-column layout does not keep intact, fall
  here. The count is printed so the limit is visible rather than implied.

```{r paper}
## helpers for the derived tier -------------------------------------------------
nums <- function(p, x) {
  m <- regmatches(x, gregexpr(p, x))[[1]]
  as.numeric(sub("^[^0-9]*([0-9.]+).*", "\\1", m))
}
sumall <- function(pats, x) sum(unlist(lapply(pats, nums, x = x)))
## the TyG sentence reports III then IV; pull both triples out of it by position,
## so the displayed interval is rebuilt from the article rather than trusted
tyg <- function(x, which) {
  i <- regexpr("to define steatosis showed consistent results", x, fixed = TRUE)
  if (i < 0) return(NA_character_)
  seg <- substr(x, i, i + 400)
  m <- regmatches(seg, gregexpr("[0-9.]+, 95%CI [0-9.]+ to [0-9.]+", seg))[[1]]
  if (length(m) < which) return(NA_character_)
  v <- as.numeric(regmatches(m[which], gregexpr("[0-9.]+", m[which]))[[1]])
  sprintf("%.1f (%.1f–%.1f)", v[1], v[3], v[4])   # v[2] is the 95 of "95%CI"
}

PAPER <- list(
  n_approx    = list(d = "~6,300", v = "6300 adults",
                     mk = function(x) paste0("~", fmtn(nums("[0-9]{4} adults with MASLD", x)[1]))),
  groups      = list(d = "386 / 4,541 / 769 / 604",
                     v = c("386 5 (1.3)", "4541 388 (8.5)", "769", "604")),
  deaths      = list(d = "585",                       v = c("5 (1.3)","388 (8.5)",
                                                            "74 (9.6)","118 (19.5)"),
                     mk = function(x) fmtn(sumall(c("[0-9]+ \\(1\\.3\\)", "[0-9]+ \\(8\\.5\\)",
                                                    "[0-9]+ \\(9\\.6\\)", "[0-9]+ \\(19\\.5\\)"), x))),
  iv_i_unadj  = list(d = "15.13",                     v = "15.13"),
  iv_i_model2 = list(d = "2.893",                     v = "2.893"),
  t2_m1_iii   = list(d = "3.175",                     v = "3.175"),
  t2_m1_iv    = list(d = "3.155",                     v = "3.155"),
  t2_m2_iii   = list(d = "3.029",                     v = "3.029"),
  t2_m2_iv    = list(d = "2.893",                     v = "2.893"),
  tyg_iii     = list(d = "3.1 (1.3–7.7)", v = "aHR: 3.1, 95%CI 1.3 to 7.7",
                     mk = function(x) tyg(x, 1)),
  tyg_iv      = list(d = "3.1 (1.2–7.6)", v = "3.1, 95%CI 1.2 to 7.6",
                     mk = function(x) tyg(x, 2)),
  died_iv     = list(d = "19.5%",                     v = "118 (19.5)"),
  died_iii    = list(d = "9.6%",                      v = "74 (9.6)",
                     mk = function(x) paste0(nums("group III \\([0-9.]+%\\)", x)[1], "%")),
  cv_deaths   = list(d = "158",                       v = c("1 (0.3)","112 (2.5)",
                                                            "19 (2.5)","26 (4.3)"),
                     mk = function(x) fmtn(sumall(c("[0-9]+ \\(0\\.3\\)", "[0-9]+ \\(2\\.5\\)",
                                                    "[0-9]+ \\(4\\.3\\)"), x))),
  conclusion  = list(d = paste("Non-obesity MASLD with high central adiposity confers",
                               "the poorest survival and cardiometabolic burden."),
                     v = "confers the poorest survival"),
  km_claim    = list(d = "least favourable survival", v = "least favourable survival"),
  independent = list(d = "independent of total body weight",
                     v = "independent of total body weight"))

pdfp <- find_paper()
have <- !is.na(pdfp) && nzchar(Sys.which("pdftotext"))
if (have) {
  ptxt <- paste(system2("pdftotext", c("-layout", shQuote(pdfp), "-"), stdout = TRUE),
                collapse = " ")
  ptxt <- gsub("\\s+", " ", gsub("[^ -~]", " ", ptxt, useBytes = TRUE), useBytes = TRUE)
}
chk <- do.call(rbind, lapply(names(PAPER), function(k) {
  e  <- PAPER[[k]]
  tr <- if (!is.null(e$mk)) "derived" else if (e$d %in% e$v) "exact" else "presence"
  found <- if (have) all(vapply(e$v, grepl, logical(1), x = ptxt, fixed = TRUE)) else NA
  der <- NA_character_
  if (isTRUE(found) && tr == "derived") {
    der <- tryCatch(e$mk(ptxt), error = function(z) NA_character_)
    if (is.na(der) || !identical(der, e$d)) found <- FALSE   # our display is wrong
  }
  data.frame(fact = k, tier = tr, value = e$d, rebuilt_from_pdf = der,
             checked_for = paste(e$v, collapse = " + "), found = found)
}))
if (have && any(!chk$found)) {
  print(chk[!chk$found, ])
  stop("a figure attributed to the paper is not supported by the paper")
}
if (have) {
  n <- table(factor(chk$tier, levels = c("exact","derived","presence")))
  cat(sprintf("%d quoted figures checked against %s: %d exact, %d rebuilt from the PDF, %d presence-only\n",
              nrow(chk), basename(pdfp), n[["exact"]], n[["derived"]], n[["presence"]]))
  cat(sprintf("The %d presence-only facts are confirmed to occur but their displayed form is not re-derived: %s\n",
              n[["presence"]], paste(chk$fact[chk$tier == "presence"], collapse = ", ")))
} else cat("paper PDF unavailable here - quoted figures NOT verified in this render\n")

emit_results("paper", setNames(lapply(PAPER, `[[`, "d"), names(PAPER)),
             list(verification = chk))
chk[, c("fact", "tier", "value", "rebuilt_from_pdf", "found")]
```

## M1's own analysis 1 — the replication

The paper's own parameterization, on the **full analytic file**. Kept separate from
the locked domain used for the diagnostics: mixing row sets in one column is how a
replication table stops being a replication table.

```{r replication}
d <- readRDS(file.path(repro, "data/derived/masld_analytic.rds"))
d$smk <- ifelse(d$SMQ020 == 1 & d$SMQ040 %in% c(1, 2), 1L, ifelse(is.na(d$SMQ020), NA, 0L))
cov1 <- c("age","female","dyslip","htn","t2dm","ckd","mi","cancer",
          "on_htn_med","on_dm_med","on_lipid_med")
fit <- function(covs = NULL)
  coxph(as.formula(paste("Surv(time_yr,dead) ~", paste(c("group", covs), collapse = "+"))), d)
m0 <- fit(); m1 <- fit(cov1); m2 <- fit(c(cov1, "smk", "sedentary_min"))
pt <- function(f, dp) sprintf(paste0("%.", dp, "f"), summary(f)$conf.int["groupIV", 1])

rbind(Unadjusted = sapply(c("groupII","groupIII","groupIV"), function(g) ci_of(m0, g)),
      `Model 1`  = sapply(c("groupII","groupIII","groupIV"), function(g) ci_of(m1, g)),
      `Model 2`  = sapply(c("groupII","groupIII","groupIV"), function(g) ci_of(m2, g)))
```

**Cause-of-death coding.** The paper's cardiovascular analysis counts diseases of
the heart only; our `cv_death` indicator also includes cerebrovascular deaths. Both
are reported so the discrepancy is a documented choice, not a silent one.

```{r cvdeath}
heart  <- sum(d$ucod_leading == 1, na.rm = TRUE)
cerebr <- sum(d$ucod_leading == 5, na.rm = TRUE)
data.frame(definition = c("diseases of heart only", "heart + cerebrovascular"),
           n = c(heart, heart + cerebr),
           note = c("matches the paper's CV total", "our pipeline's cv_death"))
```

## M1's own analysis 2 — the design-aware interaction

P3's deferred question, answered with P4's rule: build the design on the whole
fasting frame, then reach the Groups III–IV rows with `subset()`.

```{r interaction}
frame <- readRDS(file.path(repro, "data/derived/merged_all.rds"))
frame$in_analysis <- frame$SEQN %in% d$SEQN
frame$group <- factor(frame$group, levels = c("I","II","III","IV"))
fast <- frame %>% filter(!is.na(WTSAF2YR), WTSAF2YR > 0) %>% mutate(wt = WTSAF2YR / 6)
COVns <- c("age","dyslip","htn","t2dm","ckd","mi","cancer",
           "on_htn_med","on_dm_med","on_lipid_med")
fast$IV   <- as.integer(fast$group == "IV")
fast$keep <- fast$in_analysis & fast$group %in% c("III","IV") &
             complete.cases(fast[, c(COVns, "female")])

fw <- svycoxph(as.formula(paste("Surv(time_yr,dead) ~ IV*female +",
                                paste(COVns, collapse = " + "))),
               design = subset(svydesign(ids = ~SDMVPSU, strata = ~SDMVSTRA,
                                         weights = ~wt, data = fast, nest = TRUE), keep))
b  <- coef(fw)[["IV:female"]]
se <- sqrt(vcov(fw)["IV:female", "IV:female"])
data.frame(Model = c("unweighted, Model 1 (P3)", "design-aware, Model 1 (M1)"),
           rHR   = c(R$P3$facts[["rhr_unweighted"]], ci(exp(b), exp(b - 1.96*se), exp(b + 1.96*se))),
           p     = c(R$P3$facts[["p_unweighted"]],   sprintf("%.3f", 2 * pnorm(-abs(b / se)))),
           rows  = c(R$P3$facts[["n"]], fmtn(sum(fast$keep))))
```

The row count must match P3's, or the two rows are not comparable:
`r sprintf("M1 used %s rows; P3 used %s.", fmtn(sum(fast$keep)), R$P3$facts[["n"]])`

## What M1 contributes, emitted as facts

```{r emit}
ratio <- function(a, b) sprintf("%.2f", as.numeric(a) / as.numeric(b))
## point estimate out of a formatted "est (lo-hi)" fact, so the TyG ratio is derived
## from the same verified strings as the Table-2 ratios rather than asserted
est <- function(s) as.numeric(sub("^\\s*([0-9.]+).*", "\\1", s))
pf <- lapply(PAPER, `[[`, "d")
emit_results("M1", list(
  iv_i_unadj      = pt(m0, 2),
  iv_i_unadj_ci   = ci_of(m0),
  iv_i_model2     = pt(m2, 3),
  iv_i_model2_ci  = ci_of(m2),
  cv_heart        = fmtn(heart),
  cv_heart_cerebr = fmtn(heart + cerebr),
  rhr_design      = ci(exp(b), exp(b - 1.96*se), exp(b + 1.96*se)),
  p_design        = sprintf("%.3f", 2 * pnorm(-abs(b / se))),
  n_design        = fmtn(sum(fast$keep)),
  # the IV-vs-III contrast implied by the paper's own published point estimates
  paper_iv_iii_m1 = ratio(pf$t2_m1_iv, pf$t2_m1_iii),
  paper_iv_iii_m2 = ratio(pf$t2_m2_iv, pf$t2_m2_iii),
  # the TyG estimates are reported to one decimal; carry that precision through
  paper_iv_iii_tyg = sprintf("%.1f", est(pf$tyg_iv) / est(pf$tyg_iii))
))
```

`paper_iv_iii_tyg` is **computed from the two TyG point estimates**, which are
themselves rebuilt from the article's own sentence and compared against what we
transcribed. It was previously the literal string "1.0" — a small hole in the
architecture, since a change in either verified estimate would not have moved it. It
comes out at 1.0 because the article reports Group III and Group IV at the same rounded
aHR against Group I. It is a point-estimate observation, and the uncertainty around it
cannot be recovered from two separately reported intervals.

## The assembled tables

Everything below is read from the increments. No model is fitted here.

```{r assembled-funnel}
R$P1$funnel
```

```{r assembled-matrix}
R$P4$matrix
```

```{r assembled-p2}
R$P2$estimands
```

```{r assembled-p3}
R$P3$additive
```

```{r assembled-p5}
R$P5$locked_estimand
```

## The fact index

Every value the deck may cite, with the exact reference that resolves to it. This
is the provenance table: a slide showing `1.29` shows it because
`{{ P4.facts.w_m1_iii }}` resolved to it here.

```{r index}
idx <- do.call(rbind, lapply(sprintf("P%d", 1:5), function(n) {
  f <- read_results(n)$facts
  data.frame(reference = sprintf("{{ %s.facts.%s }}", n, names(f)),
             value = unlist(f, use.names = FALSE))
}))
idx <- rbind(idx, do.call(rbind, lapply(c("M1", "paper"), function(n) {
  f <- read_results(n)$facts
  data.frame(reference = sprintf("{{ %s.facts.%s }}", n, names(f)),
             value = substr(unlist(f, use.names = FALSE), 1, 70))
})))
cat(nrow(idx), "citable facts\n")
idx
```
