---
title: "P1 — Analytic cohort: step-by-step reproducible code"
subtitle: "SPPH 604 · demonstration paper (Kueh et al., BMJ Open 2026;16:e113719)"
format:
  html:
    embed-resources: true
    toc: true
    code-tools: true
    df-print: kable
execute:
  warning: false
  message: false
---

> **What this file does.** It re-derives the eligibility funnel and the four
> phenotype groups from the merged NHANES data and shows that we reach
> **N = 6,371** and groups **391 / 4585 / 787 / 608**. The full multi-cycle
> download + merge is in `../../reproduction/R/01_download.R` and
> `02_build_analytic.R`; here we read that merged frame and expose the
> eligibility logic itself, because *deciding who is in the sample* is the P1 skill.

```{r setup}
source("../_shared/paths.R")   # find_repro(), emit_results(), ci(), fmtn(), endash()
repro <- find_repro()
suppressPackageStartupMessages({library(dplyr)})
# Merged, all-participant frame produced by 02_build_analytic.R (raw + derived cols)
d <- readRDS(file.path(repro, "data/derived/merged_all.rds"))
yn <- function(x) as.integer(x == 1)
nrow(d)   # all NHANES 2007-2018 records
```

## Step 1 — Fatty Liver Index and steatosis

FLI (Bedogni 2006) needs fasting triglycerides, GGT, BMI and waist. Steatosis = FLI ≥ 60.

```{r fli}
d <- d %>% mutate(
  fli = { L <- 0.953*log(LBXTR) + 0.139*BMXBMI + 0.718*log(LBXSGTSI) +
               0.053*BMXWAIST - 15.745; exp(L)/(1+exp(L))*100 },
  steatosis = as.integer(fli >= 60))
```

## Step 2 — MASLD cardiometabolic criteria (≥ 1 required)

```{r masld}
d <- d %>% mutate(
  crit_adip = as.integer(bmi >= 25 | (female==0 & waist > 94) | (female==1 & waist > 80)),
  crit_glu  = as.integer(LBXGLU >= 100 | LBXGH >= 5.7 | yn(DIQ010)==1 |
                         yn(DIQ070)==1 | yn(DIQ050)==1),
  crit_bp   = as.integer(sbp >= 130 | dbp >= 85 | yn(BPQ050A)==1),
  crit_tg   = as.integer(LBXTR >= 150 | yn(BPQ100D)==1),
  crit_hdl  = as.integer((female==0 & LBDHDD <= 40) | (female==1 & LBDHDD < 50)),
  n_crit    = rowSums(cbind(crit_adip,crit_glu,crit_bp,crit_tg,crit_hdl), na.rm = TRUE),
  masld     = as.integer(steatosis == 1 & n_crit >= 1))
```

## Step 3 — Phenotype groups (obesity BMI ≥ 30; central WHtR ≥ 0.6)

The Table headers say "BMI < 25", but that is a **mislabel** — the non-obese
groups have BMI medians of 27.6 / 28.8. The Methods-text cut (30) is what
reproduces the group sizes.

```{r groups}
d <- d %>% mutate(
  obese   = as.integer(bmi >= 30 | (asian == 1 & bmi >= 25)),
  central = as.integer(whtr >= 0.6),
  group = factor(case_when(obese==1 & central==0 ~ "I",
                           obese==1 & central==1 ~ "II",
                           obese==0 & central==0 ~ "III",
                           obese==0 & central==1 ~ "IV"),
                 levels = c("I","II","III","IV")))
```

## Step 4 — The eligibility funnel

**Every step carries a type label.** A funnel shows N falling; it does not show *why*,
and the reasons are not interchangeable. Sorting them is the point of this step.

```{r funnel}
f <- function(cond) sum(cond, na.rm = TRUE)
inFast <- !is.na(d$WTSAF2YR) & d$WTSAF2YR > 0
funnel <- tibble::tribble(
  ~Step, ~Criterion, ~N, ~Type,
  0, "NHANES 2007-2018 records",           nrow(d),                        "-",
  1, "Age >= 18",                          f(d$age>=18),                   "ELIGIBILITY",
  2, "In fasting subsample",               f(d$age>=18 & inFast),          "DESIGN",
  3, "FLI computable (TG+GGT+BMI+waist)",  f(d$age>=18 & inFast & !is.na(d$fli)),
                                                                           "MISSINGNESS",
  4, "Steatosis (FLI >= 60)",              f(d$age>=18 & inFast & d$steatosis==1),
                                                                           "TARGET POPULATION",
  5, "MASLD (>= 1 criterion)",             f(d$age>=18 & inFast & d$masld==1),
                                                                           "TARGET POPULATION",
  6, "Phenotype group assignable",         f(d$age>=18 & inFast & d$masld==1 & !is.na(d$group)),
                                                                           "MISSINGNESS",
  7, "Mortality-eligible and linked",      f(d$age>=18 & inFast & d$masld==1 & !is.na(d$group) &
                                             d$eligstat==1 & !is.na(d$mortstat) & !is.na(d$time_yr)),
                                                                           "OUTCOME ASCERTAINMENT")
funnel$Dropped <- c(NA, -diff(funnel$N))
funnel
```

### Why the labels matter

The five categories are not five flavours of the same thing. They have different
inferential consequences, and a funnel that hides the difference invites the wrong fix.

- **ELIGIBILITY** is definitional. Restricting to adults does not bias anything; it
  states who the study is about.
- **DESIGN** is the fasting subsample. This is *not* missing data. NHANES selected it
  deliberately and supplies `WTSAF2YR` to account for the extra selection stage and its
  nonresponse. The correct response is the right weight and a domain analysis, not
  imputation. Inference stays conditional on the weighting assumptions.
- **MISSINGNESS** is item nonresponse among people who *were* selected: they should have
  a value and do not. This is the only kind of step that a missing-data method can
  address.
- **TARGET POPULATION** is disease definition. People without steatosis are not missing —
  they are not who the paper is about. Counting them as attrition inflates the apparent
  missing-data problem, which is exactly the error to avoid here: the drop from step 3 to
  step 5 is the largest in the funnel and **none of it is missingness**.
- **OUTCOME ASCERTAINMENT** is failure to link to mortality follow-up. It looks like
  missingness but attaches to the outcome rather than a covariate, so it threatens the
  analysis differently and is worth separating.

The single most common error on a cohort funnel is to call every downward step
"attrition" and then reach for imputation. Three of these five categories cannot be
imputed, and one of them should not be.

## Step 5 — Reproduced group sizes and deaths (vs paper)

```{r groups-check}
coh <- d %>% filter(age>=18, masld==1, !is.na(group), eligstat==1,
                    !is.na(mortstat), !is.na(time_yr))
data.frame(
  Group  = c("I","II","III","IV"),
  Ours_N = as.integer(table(coh$group)),
  Paper_N= c(386,4541,769,604),
  Ours_deaths = as.integer(tapply(coh$dead, coh$group, sum, na.rm=TRUE)),
  Paper_deaths= c(5,388,74,118))
cat("Total N:", nrow(coh), "(paper 6300)   Total deaths:",
    sum(coh$dead,na.rm=TRUE), "(paper 585)\n")
```

## Step 6 — The alcohol operationalization (documented ambiguity)

One sentence in the Methods, coded three defensible ways, spans > 2,000 people.
We do **not** tune to hit N; we keep the no-exclusion cohort (which matches the
paper) and flag the discrepancy.

```{r alcohol}
base <- coh
## NHANES codes "refused" as 777 and "don't know" as 999. Left raw, a rule like
## `ALQ130 > 2` silently reclassifies every refusal as a heavy drinker: here that
## is 9 people, which moves the excluded count from 2,063 to 2,072. Reserved
## codes are not measurements, and every increment must strip them before use.
base$alq <- ifelse(base$ALQ130 %in% c(777, 999), NA, base$ALQ130)
crude <- with(base, (female==0 & alq>2) | (female==1 & alq>1))            # ALQ130 alone
avg   <- with(base, (female==0 & avg_drinks_day>2) | (female==1 & avg_drinks_day>1))
data.frame(
  coding = c("Crude ALQ130 rule", "Frequency-weighted average", "No effective exclusion"),
  excluded = c(sum(crude,na.rm=TRUE), sum(avg,na.rm=TRUE), 0),
  N_remaining = c(nrow(base)-sum(crude,na.rm=TRUE), nrow(base)-sum(avg,na.rm=TRUE), nrow(base)))
```

## Step 7 — Residual N gap (honest accounting)

```{r ngap}
cc <- coh %>% filter(!is.na(cancer), !is.na(stroke), !is.na(hf), !is.na(LBXSCR))
cat("Full cohort:", nrow(coh), " | complete-case on covariates:", nrow(cc),
    " | missing serum creatinine:", sum(is.na(coh$LBXSCR)), "\n")
```

## Results for M1

M1 cites these by name; it does not re-run this file. Every number is formatted
once, here, so the deck cannot reformat or mistype it.

```{r emit}
fn <- funnel$N; names(fn) <- paste0("s", funnel$Step)
emit_results("P1", list(
  full_n   = fmtn(nrow(coh)),
  groups   = paste(fmtn(as.integer(table(coh$group))), collapse = " / "),
  deaths   = fmtn(sum(coh$dead, na.rm = TRUE)),
  locked_n = fmtn(sum(!is.na(coh$WTSAF2YR) & coh$WTSAF2YR > 0)),
  funnel_records   = fmtn(fn[["s0"]]),
  funnel_adults    = fmtn(fn[["s1"]]),
  funnel_fasting   = fmtn(fn[["s2"]]),
  funnel_fli       = fmtn(fn[["s3"]]),
  funnel_steatosis = fmtn(fn[["s4"]]),
  funnel_masld     = fmtn(fn[["s5"]]),
  funnel_linked    = fmtn(fn[["s7"]]),
  item_nonresponse = fmtn(funnel$Dropped[funnel$Step == 3]),
  outcome_loss     = fmtn(funnel$Dropped[funnel$Step == 7]),
  # what "read every downward step as attrition" would actually total
  all_downward_steps = fmtn(fn[["s1"]] - fn[["s7"]]),
  alq_excluded = fmtn(sum(crude, na.rm = TRUE)),
  alq_n        = fmtn(nrow(base) - sum(crude, na.rm = TRUE)),
  avg_excluded = fmtn(sum(avg, na.rm = TRUE)),
  avg_n        = fmtn(nrow(base) - sum(avg, na.rm = TRUE)),
  alq_reserved = fmtn(sum(base$ALQ130 %in% c(777, 999)))
), list(funnel = as.data.frame(funnel)))
```
