---
title: "P5 — Missing data: step-by-step reproducible code"
subtitle: "SPPH 604 · demonstration paper (Kueh et al., BMJ Open 2026;16:e113719)"
format:
  html:
    embed-resources: true
    toc: true
    code-tools: true
    df-print: kable
execute:
  warning: false
  message: false
---

> Starts from **exactly** the target population and survey design used for the
> corrected main estimate in P4, and uses **exactly** the paper's Model 1
> covariates. No variable is added to the model in order to create a
> missing-data problem.

```{r setup}
source("../_shared/paths.R")   # find_repro(), emit_results(), ci(), fmtn(), endash()
repro <- find_repro()
suppressPackageStartupMessages({library(dplyr); library(survey); library(survival)
                                library(mice); library(mitools)})
options(survey.lonely.psu = "adjust")

m <- readRDS(file.path(repro,"data/derived/merged_all.rds"))
a <- readRDS(file.path(repro,"data/derived/masld_analytic.rds"))
m$in_analysis <- m$SEQN %in% a$SEQN
m$group <- factor(m$group, levels=c("I","II","III","IV"))
fast <- m %>% filter(!is.na(WTSAF2YR), WTSAF2YR > 0) %>% mutate(wt = WTSAF2YR/6)
dom  <- fast %>% filter(in_analysis)

## The paper's Model 1. Nothing else.
COV <- c("age","female","dyslip","htn","t2dm","ckd","mi","cancer",
         "on_htn_med","on_dm_med","on_lipid_med")
f1  <- as.formula(paste("Surv(time_yr,dead) ~ group +", paste(COV, collapse=" + ")))
gv  <- function(f,t="groupIV"){s<-summary(f); c(s$conf.int[t,1],s$conf.int[t,3],s$conf.int[t,4])}
fm  <- function(x) sprintf("%.2f (%.2f-%.2f)", x[1],x[2],x[3])
```

## Step 0 — Three different kinds of downward step

A cohort funnel shows N falling. It does **not** show *why*, and the reasons have
completely different inferential meanings. Sort every step into one of three bins
before you call anything "missing data".

```{r waterfall}
w <- function(lbl, df, kind) cat(sprintf("%-38s %6d  %13s   %s\n", lbl, nrow(df),
        format(round(sum(df$wt)), big.mark=","), kind))
ad  <- fast %>% filter(!is.na(age), age>=18)
fli <- ad   %>% filter(!is.na(fli))
st  <- fli  %>% filter(fli>=60)
cat(sprintf("%-38s %6s  %13s   %s\n","step","n","weighted pop","kind"))
w("fasting subsample frame",      fast, "DESIGN     - the weight handles this")
w("  adults 18+",                 ad,   "ELIGIBILITY- definitional")
w("  FLI computable",             fli,  "MISSINGNESS- item nonresponse")
w("  steatosis, FLI >= 60",       st,   "DISEASE    - target population")
w("  MASLD, linked, group known", dom,  "DISEASE + small missingness")
```

- **Design.** Fasting measures come from a *designed* NHANES subsample. Its absence
  elsewhere is not missing data; it is why `WTSAF2YR` exists.
- **Disease.** Most of the drop is people who do not have steatosis. They are not
  missing — they are *not in the target population*.
- **Missingness.** Only the item-nonresponse step is a missing-data problem, and it
  is `r sprintf("%.1f%%", 100*mean(is.na(ad$fli)))` of adults in the fasting frame.

The single most common error on a funnel is to call every step "attrition".

## Step 1 — The missingness pattern, before choosing any method

```{r pattern}
miss <- sapply(COV, function(v) sum(is.na(dom[[v]])))
for (v in names(miss)[miss>0])
  cat(sprintf("  %-14s missing %4d   %.2f%% unweighted   %.2f%% weighted\n", v, miss[v],
      100*miss[v]/nrow(dom), 100*sum(dom$wt[is.na(dom[[v]])])/sum(dom$wt)))
cc <- complete.cases(dom[,COV])
cat(sprintf("\n  variables with any missingness: %d of %d\n", sum(miss>0), length(COV)))
cat(sprintf("  complete cases: %d of %d  (loses %d; %.2f%% unweighted, %.2f%% weighted)\n",
    sum(cc), nrow(dom), sum(!cc), 100*mean(!cc), 100*sum(dom$wt[!cc])/sum(dom$wt)))
```

**One** Model-1 variable carries any missingness at all. There is no overlap pattern
to examine, because there is nothing to overlap with.

## Step 2 — State the estimand before comparing anything

> The **weighted, Model-1-adjusted hazard ratio for Group IV versus Group I**, in the
> MASLD domain of the NHANES fasting subsample, 2007–2018.

Fixing this first matters: complete-case analysis and multiple imputation can
silently target *different populations*, and then a difference between them is not
evidence about missing data.

## Step 3 — Complete case, on the corrected design

```{r cc}
des <- subset(svydesign(ids=~SDMVPSU, strata=~SDMVSTRA, weights=~wt,
                        data=fast, nest=TRUE), in_analysis)
fit_w  <- svycoxph(f1, design=des)
fit_u  <- coxph(f1, data=dom)
cat(sprintf("weighted   Model-1, complete case:  IV = %s   (n = %d)\n", fm(gv(fit_w)), fit_w$n))
cat(sprintf("unweighted Model-1, complete case:  IV = %s   (n = %d)\n", fm(gv(fit_u)), fit_u$n))
```

## Step 3b — A stress test that needs no model

The only incomplete Model-1 variable is **binary**. That permits a sensitivity
analysis a reader can check by eye: assign every one of the 109 missing values to one
extreme, then the other, and refit. This does not identify the missing values and it
is **not** a formal bounds argument — it is a transparent stress test spanning the two extreme
assignments the binary indicator permits.

```{r stress}
fitv <- function(val, lbl){
  fk <- fast
  if (!is.na(val)) fk$cancer[fk$in_analysis & is.na(fk$cancer)] <- val
  f <- svycoxph(f1, design = subset(svydesign(ids=~SDMVPSU, strata=~SDMVSTRA,
                  weights=~wt, data=fk, nest=TRUE), in_analysis))
  g <- gv(f)
  data.frame(scenario = lbl, HR = round(g[1],2),
             lo = round(g[2],2), hi = round(g[3],2), n = f$n)
}
stress <- rbind(fitv(NA, "complete case (109 dropped)"),
                fitv(0,  "all 109 assigned NO cancer"),
                fitv(1,  "all 109 assigned CANCER"))
stress
```

The Group IV estimate spans `r sprintf("%.2f to %.2f", min(stress$HR), max(stress$HR))`
across both extremes — a total range of
`r sprintf("%.2f", max(stress$HR)-min(stress$HR))` on the hazard ratio. These are **two global extreme scenarios**, not bounds over all assignments. A
pattern of missingness correlated with phenotype and covariates could move the
estimate further than either uniform assignment does, so we do not claim the true
value must lie inside this range — only that two transparent extremes leave it
essentially unchanged.

## Step 4 — Multiple imputation, survey-aware

Assumptions that would make this informative: the missingness is **MAR given the
Model-1 covariates and the outcome**, and the imputation model is correctly
specified. The outcome enters as the event indicator plus a Nelson–Aalen
cumulative-hazard term — omitting it biases covariate–outcome associations toward
the null.

**This is not fully survey-modelled multiple imputation, and we do not claim it is.**
The imputation model is a standard `mice()` fit with the sampling weight and stratum
entered *as predictors*; the design is reintroduced at the analysis stage, when each
completed dataset is refitted with `svycoxph()` on a properly subsetted design. A
fully design-aware imputation would model the sampling structure inside the
imputation itself. We report this as a **pragmatic sensitivity analysis** and check
below whether the shortcut matters here.

```{r mi}
idat <- dom[, c(COV,"group","dead","time_yr","wt","SDMVSTRA")]
idat$na_cumhaz <- mice::nelsonaalen(idat, time_yr, dead)
ini <- mice(idat, maxit=0, printFlag=FALSE); pr <- ini$predictorMatrix
pr[,"time_yr"] <- 0                                        # raw time: not a predictor
pr[c("time_yr","dead","na_cumhaz","wt","SDMVSTRA"),] <- 0  # complete: never imputed
imp <- mice(idat, m=10, predictorMatrix=pr, printFlag=FALSE, seed=604)

fits <- lapply(1:10, function(i){
  fk <- fast; fk[fk$in_analysis, COV] <- complete(imp,i)[,COV]
  svycoxph(f1, design=subset(svydesign(ids=~SDMVPSU,strata=~SDMVSTRA,weights=~wt,
                                       data=fk,nest=TRUE), in_analysis))
})
po <- summary(MIcombine(fits))["groupIV",]
cat(sprintf("weighted   Model-1, MI (m=10):      IV = %.2f (%.2f-%.2f)\n",
    exp(po[["results"]]), exp(po[["(lower"]]), exp(po[["upper)"]])))
cat(sprintf("fraction of missing information:    %s\n", po[["missInfo"]]))
```

## Step 4b — The same four approaches, on the locked estimand

Everything above is reported against **Group I**, the paper's reference, because that is
the contrast the published analysis reports. The term's locked question is **IV vs III**.
Re-levelling inside the same model gives it, and the missingness has to be checked there
too rather than assumed to behave the same way.

```{r locked}
rl3  <- function(x){ x$group <- relevel(x$group, "III"); x }
desL <- function(fk) subset(svydesign(ids=~SDMVPSU, strata=~SDMVSTRA, weights=~wt,
                                      data = rl3(fk), nest = TRUE), in_analysis)
fitL <- function(val, lbl){
  fk <- fast
  if (!is.na(val)) fk$cancer[fk$in_analysis & is.na(fk$cancer)] <- val
  f <- svycoxph(f1, design = desL(fk))
  data.frame(scenario = lbl, HR = fm(gv(f)), n = f$n)
}
fitsL <- lapply(1:10, function(i){
  fk <- fast; fk[fk$in_analysis, COV] <- complete(imp, i)[, COV]
  svycoxph(f1, design = desL(fk))
})
poL <- summary(MIcombine(fitsL))["groupIV", ]
locked <- rbind(
  fitL(NA, "complete case (109 dropped)"),
  fitL(0,  "all 109 assigned NO cancer"),
  fitL(1,  "all 109 assigned CANCER"),
  data.frame(scenario = "MI (m = 10)", n = fitsL[[1]]$n,
             HR = sprintf("%.2f (%.2f-%.2f)", exp(poL[["results"]]),
                          exp(poL[["(lower"]]), exp(poL[["upper)"]]))))
locked[, c("scenario","HR","n")]
```

The locked contrast behaves exactly as the replication contrast does: every approach
returns the same estimate to two decimals. Note that this is a statement about *this*
missingness, not a general result — it had to be computed on the locked contrast to be
claimed about the locked contrast.

## Step 5 — What this actually shows

```{r summary}
cc_w <- gv(fit_w); mi_w <- c(exp(po[["results"]]), exp(po[["(lower"]]), exp(po[["upper)"]]))
cat(sprintf("  complete case (weighted): %s\n  imputed       (weighted): %s\n  difference in HR: %.3f\n",
    fm(cc_w), fm(mi_w), mi_w[1]-cc_w[1]))
```

```{r between}
co <- sapply(fits, function(f) coef(f)[["groupIV"]])
cat(sprintf("  between-imputation variance of the Group IV log-HR: %.3e
", var(co)))
cat(sprintf("  distinct fitted values across m=10: %d
", length(unique(round(co,10)))))
```

**The missingness is real and correcting it changes essentially nothing.** One
covariate, `r sprintf("%.2f%%", 100*mean(!cc))` of the sample, and a hazard ratio that
moves in the third significant figure.

Be exact about the fraction of missing information: it prints as `r po[["missInfo"]]`,
but that is **approximately zero at the displayed precision, not literally zero**. The
between-imputation variance is `r sprintf("%.1e", var(co))` — the imputed `cancer` values really do
differ between completions, they simply have almost no leverage on the Group IV
coefficient. Reporting "0%" without that qualification would claim more than the
object supports.

It also means the shortcut flagged above is **immaterial here**: entering the weight
and stratum as predictors rather than modelling the design inside the imputation
changes the estimate not at all. That is a statement about this dataset with 1.8%
missingness on one covariate, not a general licence.

### The conclusion, stated as the evidence supports it

> Within the locked domain — the 6,048 rows P5 analyses — only 1.8% of records are
> incomplete on the paper's Model-1 covariates, entirely because of missing cancer
> status. Complete-case analysis, extreme-value stress tests, and pragmatic multiple
> imputation yield materially similar estimates. In this dataset, handling this residual
> missingness does not alter the substantive inference; accounting for the complex survey
> design has substantially greater impact.

Note the two things that sentence is careful **not** to say. It does not say
missingness is generally unimportant — it says it is numerically negligible *here*. And
it does not say survey design is the paper's dominant threat; it says design mattered
more than missingness *among the analyses run in P4 and P5*. Which threat most
endangers the paper's conclusion is a judgment that needs all five increments, and it
is made at M1, not here.

**Methodological importance is measured, not presumed.** A sensitivity analysis that
returns "no material change" has done its job.

**What *does* move the conclusion is the design, not the missing data.** Compare the
two complete-case rows in Step 3: unweighted, the adjusted interval excludes 1;
correctly weighted, it does not. Respecting the sampling design changes what you can
claim; imputing 1.8% of one covariate does not.

**A sensitivity analysis we deliberately did not run.** Adding income (PIR, ~9%
missing) and systolic BP would raise complete-case loss to about 13% and make the
missing-data exercise look substantial. Neither is in the paper's Model 1. Adding
variables to manufacture a missing-data problem changes the adjustment set and
answers a different question. If you want them, introduce them as a separately
labelled sensitivity model with its own estimand — not inside this comparison.

## Results for M1

```{r emit}
lk <- function(s) endash(locked$HR[locked$scenario == s])
emit_results("P5", list(
  n_missing   = fmtn(sum(is.na(dom$cancer))),
  pct_uw      = sprintf("%.2f", 100 * mean(!cc)),
  pct_w       = sprintf("%.2f", 100 * sum(dom$wt[!cc]) / sum(dom$wt)),
  n_vars_miss = fmtn(sum(miss > 0)),
  n_vars      = fmtn(length(COV)),
  locked_cc   = lk("complete case (109 dropped)"),
  locked_no   = lk("all 109 assigned NO cancer"),
  locked_yes  = lk("all 109 assigned CANCER"),
  locked_mi   = lk("MI (m = 10)"),
  n_cc        = fmtn(locked$n[locked$scenario == "complete case (109 dropped)"]),
  n_full      = fmtn(locked$n[locked$scenario == "MI (m = 10)"])
), list(locked_estimand = locked[, c("scenario","HR","n")],
        replication_estimand = data.frame(
          scenario = stress$scenario,
          HR = endash(sprintf("%.2f (%.2f-%.2f)", stress$HR, stress$lo, stress$hi)),
          n = stress$n)))
```
